---
title: "Data Report - Course Project 1"
author: "Akash Suresh"
date: "July 10, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and Preprocessing Data

The above mentioned data is loaded and preprocessed in the following manner. Initially, NA's are omitted for easy calculations of the total number of steps, mean and median.

```{r data,message=F, warning=F}
## Reading data
myData <- read.csv("../activity.csv")

## Load dplyr package
library(dplyr)

## Omit rows with NAs
reqData <- filter(myData,!is.na(steps))
```

## Number of Steps per Day

Grouping the data per day and using summarise() to calculate the number of steps taken per day. Since, the rows with NA are omitted, the total for each day is a valid numeric value.

```{r daywise1,message=F, warning=F}
## Grouping data by day
groupByDay <- group_by(reqData,date)

## Number of steps per day
sumPerDay <- summarise(groupByDay,sum(steps, na.rm = TRUE))

## Print sumPerDay
library(knitr)
kable(sumPerDay, format = "markdown", col.names = c("Date", "Steps"))
```

Mean of the number of steps across all days.
```{r daywise2,message=F, warning=F}
## Mean of the total number of steps taken per day
meanPerDay <- mean(sumPerDay$`sum(steps, na.rm = TRUE)`)
meanPerDay
```

Median of the number of steps across all days.
```{r daywise3,message=F, warning=F}
## Median of the total number of steps taken per day
medianPerDay <- median(sumPerDay$`sum(steps, na.rm = TRUE)`)
medianPerDay
```

Plotting a histogram of the total number of steps taken each day.
```{r daywise4,message=F, warning=F}
## Histogram of the total number of steps taken per day
hist(sumPerDay$`sum(steps, na.rm = TRUE)`, breaks = nrow(sumPerDay), xlab = "Number of steps per day", main = "Histogram of total steps per day", col = "salmon")
```

## Average Daily Activity Pattern

This section groups activity data based on 5-minute intervals. The following plot shows  5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r intervalwise1,message=F, warning=F}
## Grouping data by interval
groupByInterval <- group_by(reqData,interval)

## Number of steps per interval across all days
sumPerInterval <- summarise(groupByInterval,sum(steps, na.rm = TRUE))

## Plot of Activity vs Interval
plot(x = sumPerInterval$interval, y = sumPerInterval$`sum(steps, na.rm = TRUE)`, xlab = "Interval", ylab = "Number of steps", main = "Activity vs Interval Plot", col = "red", type = "l")
```

The interval that contains the maximum number of steps on average across all days.
```{r intervalwise2, message=F, warning=F}
## Find max interval with max steps
orderedData <- arrange(sumPerInterval,desc(`sum(steps, na.rm = TRUE)`))
orderedData$interval[1]
```

## Imputing Missing Values

Until this secion, rows with NAs were ignored. In this section NAs will be imputed with median values. Specifically, median of the 5-minute interval will be used.

Let's use the myData variable to calculate the number of NAs in the data, since it is unprocessed. 
```{r impute1,message=F, warning=F}
## Generate boolean that gives information about NAs
naOrNot <- is.na(myData$steps)
sum(naOrNot)
```

Replacing the NAs as mentioned above.
```{r impute2,message=F, warning=F}
## Find median across all intervals for all dates
temp <- split(myData[,c(1.3)], myData$interval)
temp2 <- sapply(temp, median, na.rm = TRUE)

## Combine median information with DataFrame
imputedData <- cbind(myData, temp2)

## Impute NAs with median
imputedData <- transform(imputedData, steps = ifelse(is.na(steps), temp2, steps))
```


Plotting a histogram of the total number of steps taken each day, with NAs replaced.
```{r impute3,message=F, warning=F}
## Grouping data by day
groupByDay <- group_by(imputedData,date)

## Number of steps per day
sumPerDay <- summarise(groupByDay,sum(steps))

## Histogram of the total number of steps taken per day
hist(sumPerDay$`sum(steps)`, breaks = nrow(sumPerDay), xlab = "Number of steps per day", main = "Histogram of total steps per day", col = "salmon")
```

Mean of the number of steps across all days (with imputed data).
```{r impute4,message=F, warning=F}
## Mean of the total number of steps taken per day
meanPerDay <- mean(sumPerDay$`sum(steps)`)
meanPerDay
```

Median of the number of steps across all days (with imputed data).
```{r impute5,message=F, warning=F}
## Median of the total number of steps taken per day
medianPerDay <- median(sumPerDay$`sum(steps)`)
medianPerDay
```

As expected, after the data s imputed the mean and median values change. Since a lot of the median values are small and even 0, the mean and median have decreased now. This makes sense because alot of the newly added values (ones that were NAs) are much smaller than the original mean and median.

## Weeday vs Weekend Trend

In the below code, a new factor variable, weekday, is initialized with 0 or 1. 1 stands for weekend and 0 for weekday. Also, the new vector is combined the imputed data.

```{r Weekday1,message=F, warning=F}
## A new vector that contains weekday info
weekday <- as.factor(ifelse(weekdays(as.Date(imputedData$date))%in%c("Saturday", "Sunday"), 1, 0))

## Combine with existing data
imputedData <- cbind(imputedData, weekday)

## Group data by weekday and interval
temp3 <- group_by(imputedData, weekday, interval)

## Obtain mean for the same groups
temp4 <- summarise(temp3, mean(steps))

## Ploting the required graph
library(ggplot2)

ggplot(temp4, aes(x = interval, y = `mean(steps)`, colour = weekday)) + geom_smooth() + xlab("Interval") + ylab("Average number of steps taken")
```

As mentioned above, the weekday factor variable indicates whether the given date is a weekend or not. Weekends are identified by 1. From the above plot we see that the average trend observed in weekdays and weekends are different. The highlighted portion around the curves show the confidence interval.